A semantics of core AUTOSAR

### 2016-08-23 paper/paper.tex
Dump of current paper.

@@

The AUTOSAR standard is an open software component architecture for the automotive industry. Its main purpose is to enable interoperability of software modules among different vendors and on heterogeneous platforms, via an extensive set of standardized interfaces and libraries, and a common software development methodology.

@@

An AUTOSAR model is primarily a structure of interconnected \textbf{software components} (SWCs). Links between SWCs are called \textbf{connectors}, which attach to \textbf{ports} exposed by the SWCs. A port is either \textbf{required} or \textbf{provided}, and then classified according to what \textbf{interface} it exposes: \textbf{sender-receiver}, \textbf{client-server}, \textbf{trigger} and \textbf{mode-switch} are common interface types. A sender-receiver interface is an aggregation of one or more \textbf{data-elements}, with each element characterized by the type of data it carries and optional further communication specification details. A client-server interface likewise consists of one or more \textbf{operations}, that each give the signature of a procedure call and the possible errors that may result.

@@

To interact with ports, inter-runnable variables and exclusive areas, runnables use what is abstractly known as the \textbf{virtual function bus} (VFB). Concretely, the VFB takes the shape of a C programming interface to the \textbf{run-time environment} (RTE), custom-generated for each runnable on the basis of its access constraints. Behind the RTE, a collection of OS kernels, communication stacks and other basic software modules implement the VFB services for the platform at hand. Complete AUTOSAR designs also typically contain \textbf{ECU} and \textbf{task assignments}, which are manually built tables that control how SWCs and runnables map onto the available hardware and OS resources, respectively.

@@

\subsection{Calling a server}

@@

\subsection{Passing back a server result}

@@

Runnables with an activation state distinct from $\Tidle$ or $\Teps$ are amenable to instantiation. Axiom (49) defines this crucial step for runnables triggered by all types of events except server invocation, which is instead handled by axiom (50). Both rules have many details in common:
\begin{itemize}
\item They apply only when the runnable instantiation timer (parameter two) has reached zero.
\item They reset the timer to a model-defined \emph{minimumStartInterval} to disable further instantiation immediately after a new instance is born.
\item They are guarded by the condition that the current number of runnable instances (fourth parameter) is either zero, or the runnable has been defined to accept concurrent invocations.
\item They make the runnable keep track of the new number of instances.
\item They let the new instance inherit the address of its parent runnable.
\item They create the new instance in parallel with its parent (runnable instances is actually the only example of dynamic process creation in this calculus).
\item They initialize the new instance with code according to the model-defined implementation and an empty list of owned exclusive areas.
\end{itemize}
The differences all emanate from the server/non-server distinction:
\begin{itemize}
\item The non-server runnable instance gets a $\Tvoid$ client parameter whereas the server instance is assigned the client tuple from the head of the server's client queue.
\item The non-server runnable toggles back to the $\Tidle$ activation state, while the server just chops of the head of its client queue.
\item Only the server runnable instance is given a non-void parameter, which is taken from the same client queue head.
\end{itemize}
The transition label of axioms (49) and (59) is noteworthy because it is not matched by a corresponding hearing transition; all other process terms just ignore the information that a new runnable instance has been spawned. An example of process instantiation within a parallel context will thus merely be a trivial variant of one of the instantiation axioms, and is therefore left out.

@@

$$
\begin{array}{c}
  {\Trunnable a 0 \Tidle 0} \Opar 
  {\Tqelem b 2 \Teps} \vspace{0.6ex} \\

  \red{\hear{b}{\Tsnd {v_1} \Teps}} \\

  {\Trunnable a 0 \Tpending 0} \Opar 
  {\Tqelem b 2 {v_1}} \vspace{0.6ex} \\

  \red{\say{a}{\Tnew}} \\

  {\Trunnable a 0 \Tidle 1} \Opar 
  {\Tqelem b 2 {v_1}} \,\Opar  \\
  {\Trinst a \Tvoid \Teps {\Code}} \vspace{0.6ex} \\

  \red{\hear{b}{\Tsnd {v_2} \Teps}} \\

  {\Trunnable a 0 \Tpending 1} \Opar 
  {\Tqelem b 2 {\Tcons{v_1}{v_2}}} \,\Opar  \\
  {\Trinst a \Tvoid \Teps {\Code}} \vspace{0.6ex} \\

  \red{\say{a}{\Tnew}} \\

  {\Trunnable a 0 \Tidle 2} \Opar 
  {\Tqelem b 2 {\Tcons{v_1}{v_2}}} \,\Opar  \\
  {\Trinst a \Tvoid \Teps {\Code}} \Opar 
  {\Trinst a \Tvoid \Teps {\Code}}
\end{array}
$$

$$
\begin{array}{c}
  {\Trunnable a 0 \Tidle 0} \Opar 
  {\Tqelem b 2 \Teps} \vspace{0.6ex} \\

  \red{\hear{b}{\Tsnd {v_1} \Teps}} \\

  {\Trunnable a 0 \Tpending 0} \Opar 
  {\Tqelem b 2 {v_1}} \vspace{0.6ex} \\

  \red{\hear{b}{\Tsnd {v_2} \Teps}} \\

  {\Trunnable a 0 \Tpending 0} \Opar 
  {\Tqelem b 2 {\Tcons{v_1}{v_2}}} \vspace{0.6ex} \\

  \red{\say{a}{\Tnew}} \\

  {\Trunnable a 0 \Tidle 2} \Opar 
  {\Tqelem b 2 {\Tcons{v_1}{v_2}}} \,\Opar  \\
  {\Trinst a \Tvoid \Teps {\Code}}
\end{array}
$$

\section{Discussion / results}
\label{sec:Disc}

\subsection{Ambiguities and alternative interpretations}
\label{sec:DiscAmb}

\subsection{Clarification proposals and improvements}
\label{sec:DiscImp}

irv without initValue vs.\ delem without initValue -- ``undefined'' vs.\ NO\_DATA at initial read.

Multiple runnables triggered by same OperationInvokedEvent?

Multiple timingEvent for same runnable, possibly with same period?

Timer initial offsets

@@

>>>>

\bibliographystyle{splncs03} % Latest bibtex style for llncs

@@

The AUTOSAR standard is an open software component architecture for the automotive industry \cite{AR:Homepage}. Its main purpose is to enable interoperability of software modules among different vendors and on heterogeneous platforms, via an extensive set of standardized interfaces and libraries, and a common software development methodology.

@@

An AUTOSAR model is primarily a structure of interconnected \textbf{software components} (SWCs) \cite{AR:SWC}. Links between SWCs are called \textbf{connectors}, which attach to \textbf{ports} exposed by the SWCs. A port is either \textbf{required} or \textbf{provided}, and then classified according to what \textbf{interface} it exposes: \textbf{sender-receiver}, \textbf{client-server}, \textbf{trigger} and \textbf{mode-switch} are common interface types. A sender-receiver interface is an aggregation of one or more \textbf{data-elements}, with each element characterized by the type of data it carries and optional further communication specification details. A client-server interface likewise consists of one or more \textbf{operations}, that each give the signature of a procedure call and the possible errors that may result.

@@

To interact with ports, inter-runnable variables and exclusive areas, runnables use what is abstractly known as the \textbf{virtual function bus} (VFB). Concretely, the VFB takes the shape of a C programming interface to the \textbf{run-time environment} (RTE), custom-generated for each runnable on the basis of its access constraints \cite{AR:RTE}. Behind the RTE, a collection of OS kernels, communication stacks and other basic software modules implement the VFB services for the platform at hand. Complete AUTOSAR designs also typically contain \textbf{ECU} and \textbf{task assignments}, which are manually built tables that control how SWCs and runnables map onto the available hardware and OS resources, respectively.

@@

\subsection{Calling a server} \label{sect:Call}

@@

\subsection{Passing back a server result} \label{sect:Res}

@@

Runnables with an activation state distinct from $\Tidle$ or $\Teps$ are amenable to instantiation. Axiom (49) defines this crucial step for runnables triggered by all types of events except server invocation, which is instead handled by axiom (50). Both rules have many details in common:
\begin{itemize}
\item They apply only when the runnable instantiation timer (parameter two) has reached zero.
\item They reset the timer to a model-defined \emph{minimumStartInterval} to disable further instantiation immediately after a new instance is born.
\item They are guarded by the condition that the current number of runnable instances (fourth parameter) is either zero, or the runnable has been defined to accept concurrent invocations.
\item They make the runnable keep track of the new number of instances.
\item They let the new instance inherit the address of its parent runnable.
\item They create the new instance in parallel with its parent (runnable instances is actually the only example of dynamic process creation in this calculus).
\item They initialize the new instance with code according to the model-defined implementation and an empty list of owned exclusive areas.
\end{itemize}
The differences all emanate from the server/non-server distinction:
\begin{itemize}
\item The non-server runnable instance gets a $\Tvoid$ client parameter whereas the server instance is assigned the client tuple from the head of the server's client queue.
\item The non-server runnable toggles back to the $\Tidle$ activation state, while the server just chops of the head of its client queue.
\item Only the server runnable instance is given a non-void code parameter, which is taken from the same client queue head.
\end{itemize}
The transition label of axioms (49) and (59) is noteworthy because it is not matched by a corresponding hearing transition; all other process terms just ignore the information that a new runnable instance has been spawned. An example of process instantiation within a parallel context will thus merely be a trivial variant of one of the instantiation axioms, and is therefore left out.

@@

The transition relation defined in Appendix~\ref{sec:Rules} is non-deterministic; i.e., it may very well be that for some process $p$,
$$
\begin{array}{c}
	p \red{l_1} p_1 \hspace{10mm}\text{and}\hspace{10mm} p \red{l_2} p_2
\end{array}
$$
are both possible. If it also holds that
$$
\begin{array}{c}
	p_1 \red{l_2} q \hspace{10mm}\text{and}\hspace{10mm} p_2 \red{l_1} q
\end{array}
$$
are derivable, we note that process $p$ may evolve into state $q$ either via an $l_1$ step followed by $l_2$, or by first taking the $l_2$ step and then $l_1$. Such freedom of choice is commonly taken as the definition of {\em parallelism} in the process calculus literature --- if the sequential ordering of steps $l_1$ and $l_2$ has no significance, they might just as well be considered to occur in parallel.

However, it may also be that two possible step orderings do not lead to identical results. That happens if, instead of the two transitions leading to $q$ above, we have
$$
\begin{array}{c}
	p_1 \red{l_2} q_1 \hspace{10mm}\text{and}\hspace{10mm} p_2 \red{l_1} q_2
\end{array}
$$
where $q_1$ and $q_2$ are "similar" but not identical states. This non-determinism also has a concrete interpretation, either reflecting the non-deterministic nature of a system's environment (messages $l_1$ and $l_2$ just happen to arrive in some particular order) or its internals (processor speed or a scheduling choice just happens to pick $l_1$ before $l_2$). It is one of the primary merits of the process calculus approach to allow the scope of such inherent non-determinism to be precisely defined, because although a given system state may render many different behaviors possible, many more are not. A formal semantics makes it possible to clearly draw the line between these groups of behaviors.

The AUTOSAR specification contains a very subtle sensitivity to this kind of non-deterministism that is worth examplifying. Consider a queued data element $a$ together with a runnable $b$ set up to trigger on data reception events for $a$. A scenario where two values $v_1$ and $v_2$ are sent on a port connected to $a$ may lead to the following trace:
$$
\begin{array}{c}
  {\Tqelem a 2 \Teps} \Opar 
  {\Trunnable b 0 \Tidle 0} \vspace{0.6ex} \\

  \red{\hear{i.p.e}{\Tsnd {v_1} \Teps}} \\

  {\Tqelem a 2 {v_1}} \Opar 
  {\Trunnable b 0 \Tpending 0} \vspace{0.6ex} \\

  \red{\say{b}{\Tnew}} \\

  {\Tqelem a 2 {v_1}} \Opar 
  {\Trunnable b 0 \Tidle 1} \,\Opar  \\
  {\Trinst b \Tvoid \Teps {\Code}} \vspace{0.6ex} \\

  \red{\hear{i.p.e}{\Tsnd {v_2} \Teps}} \\

  {\Tqelem a 2 {\Tcons{v_1}{v_2}}} \Opar 
  {\Trunnable b 0 \Tpending 1} \,\Opar  \\
  {\Trinst b \Tvoid \Teps {\Code}} \vspace{0.6ex} \\

  \red{\say{b}{\Tnew}} \\

  {\Tqelem a 2 {\Tcons{v_1}{v_2}}} \Opar 
  {\Trunnable b 0 \Tidle 2} \,\Opar  \\
  {\Trinst b \Tvoid \Teps {\Code}} \Opar 
  {\Trinst b \Tvoid \Teps {\Code}}
\end{array}
$$
That is, a resulting state where $a$ holds the two received values together with two concurrent instances of $b$ ready to start reacting upon the two reception events.

However, an alternative trace is equally valid for the same scenario:
$$
\begin{array}{c}
  {\Tqelem a 2 \Teps} \Opar 
  {\Trunnable b 0 \Tidle 0} \vspace{0.6ex} \\

  \red{\hear{i.p.e}{\Tsnd {v_1} \Teps}} \\

  {\Tqelem a 2 {v_1}} \Opar 
  {\Trunnable b 0 \Tpending 0} \vspace{0.6ex} \\

  \red{\hear{i.p.e}{\Tsnd {v_2} \Teps}} \\

  {\Tqelem a 2 {\Tcons{v_1}{v_2}}} \Opar 
  {\Trunnable b 0 \Tpending 0} \vspace{0.6ex} \\

  \red{\say{b}{\Tnew}} \\

  {\Tqelem a 2 {\Tcons{v_1}{v_2}}} \Opar 
  {\Trunnable b 0 \Tidle 1} \,\Opar  \\
  {\Trinst b \Tvoid \Teps {\Code}}
\end{array}
$$
Here there is only one $b$ instance created and none is pending, even though $a$ correctly contains the two values received. The reason is that by swapping the order of the first instantiation and the second data reception steps, the system is put in a state where its inability to count pending instantiations becomes fatal.

Unfortunately there is very little a programmer can do to influence which of these two traces is taken. Giving $b$ a positive {\em minimumStartInterval} just guarantees that any subsequent events are lost during that interval. Long delays between reception events will reduce the likelihood of losses, but the choice of when to take an instantiation step is still fully internal to the operating system scheduler. Moreover, when messages are produced by multiple parallel senders, maintaining control over their time distribution may be highly impractical. Non-concurrent runnables can work around the problem by iterating over all received data at each invocation, but that solution is not very attractive if concurrent processing of each received item is desired.

One could argue that replacing the {\bf pending}/{\bf idle} flag of runnables with a small counter variable would render the event triggering mechanism in AUTOSAR much more robust. The standard explicitly excludes event counting for all events but operation invocation, though, so the possibility that events may be lost is undeniably intentional \cite[ch.~4.2.3]{AR:RTE}. However, should this design decision ever be revisited, the two trace alternatives above could serve as a succinct problem characterization.

\section{Ambiguities and clarification proposals}
\label{sec:Amb}

The formal semantics presented in this report is aimed to be a faithful capture of the AUTOSAR specification, especially its Software Components template and RTE specification \cite{AR:SWC, AR:RTE}. Still, there are several areas where we have found the specification open to interpretation, or at least in need of some further clarification. We detail a few of these areas here, as an illustration to the kind of design decisions that may surface as the result of a formalization undertaking.

Unqueued data elements that are read before first written should return the error code {\bf neverReceived} \cite[ch.~5.6.10]{AR:RTE}. Since no exceptions are mentioned in the specification, our intepretation is that this holds even when a data element possesses an {\em initValue} attribute  (third parameter of the initial {\bf delem} state defined in Section~\ref{sec:Sem}). However, as this renders any such assignments entirely redundant, an alternative interpretation creates a term
$$
  \Tdelem {i.p.e} \Tfalse v
$$
for each unqueued data element $i.p.e$, where $v$ is the {\em initValue} of $e$ if it exists, and $\TneverReceived$ otherwise.

Inter-runnable variables can also be read before written, although here there is no special error code defined in the specification for the uninitialized case \cite[ch.~5.6.26]{AR:RTE}. Section~\ref{sec:Sem} sets each {\bf irv} contents to {\bf void} to indicate a truly non-existing value, although an entirely reasonable alternative would be
$$
  \Tirv {i.s} v
$$
where $v$ is the {\em initValue} of $s$ if it exists, and the error code {\bf nodata} otherwise.

The specification restricts client-server connections such that a client may not be connected to multiple servers, for the purpose of avoiding an operation call to be handled by more than one server runnable \cite[ch.~4.2.3]{AR:RTE}. However, this restriction does not explicitly forbid multiple {\em OperationInvokedEvent}s refering to the same client-server operation to trigger separate runnables. Our formulation of client-server communication is based on the assumption that this is just an oversight, as allowing multiple events would have the same problem with ambiguous response values as a would one-to-many client-server connection pattern. %It is not easy to see what could constitute an alternative interpretation without introducing some programmer option for selecting among server results.

Although the AUTOSAR specification only motivates timed runnable activation with the need to support plain periodic execution, there is really nothing that precludes such runnables from being triggered by additional events as well --- including other {\em TimingEvent}s, that might even carry identical periods. Our formulation naturally supports this generality, but it should be observed that the inability of AUTOSAR runnables to count pending activations (see Section~\ref{sec:Parallelism}) also renders the resolution of any coinciding timer events highly non-deterministic. Restricting

It may further be noted that an alternative interpretation where conflicting timers are outruled just means a static model restriction, with no semantic implications on the timer configurations that are actually allowed.

Related to multiple timers is the

@@

\bibliography{refs}
